---
title: "Communicable Diseases in Toronto: Trends and Implications for Prevention^[All code and data are available at: https://github.com/CampbelSoup/Toronto-Communicable-Disease-Report.]"
author: "Joshua Campbell"
date: "2026-09-28"
date-format: "D MMMM YYYY"
bibliography: references.bib
format:
  pdf:
    keep-tex: true
    fig-pos: H
    tbl-pos: H
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{table}{H}
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
#| echo: false
# Code for bringing in the data to that I don't have to reread it for every figure.

library(readr)
library(dplyr)
library(ggplot2)
library(tinytable)

# Keep PDF tables at their position in the document.
options(tinytable_latex_placement = "H")

disease_data <- read_csv(
  "../data/processed/yearly_cases_2021_2025.csv",
  col_types = cols(`Yearly Rate` = col_character()),
  show_col_types = FALSE
)

vaccine_data <- disease_data |>
  filter(`Disease Category` == "Vaccine Preventable Diseases")
```


Communicable disease surveillance programs help public health agencies understand how infectious disease patterns change over time and provide evidence for targeted population health initiatives. This report analyzes Toronto’s Monthly Communicable Disease Surveillance Data between 2021 and 2025 using monthly case totals across six disease categories and population-adjusted rates to examine seasonal patterns. Vaccine-preventable diseases showed clear winter seasonality, with influenza accounting for 81.4% to 94.1% of reported cases in the category between 2022 and 2025 and reaching an annual rate of 387.0 cases per 100,000 in 2025. These findings highlight influenza as an important target for seasonal prevention efforts and show how surveillance data can be used alongside vaccination coverage to provide evidence in support of public health programming.  


# 1. Introduction
Communicable disease surveillance plays an important role in understanding how infectious diseases affect a population over time. By tracking reported cases across diseases and time periods, public health agencies can identify emerging patterns and support decision-making related to prevention and population health [@HSP_2026]. In Toronto, Toronto Public Health (TPH) publishes a monthly surveillance dataset describing reported cases of a wide range of communicable diseases [@torontoCommunicableData]. This dataset is used to analyze how patterns of illness have changed in the city over recent years.

This report examines Toronto’s Monthly Communicable Disease Surveillance Data between 2021 and 2025, excluding COVID-19. The report considers monthly totals across six broad disease categories to observe prominent seasonal trends and overall volume of reported cases. Particular attention is given to categories with clear seasonal variation, notably vaccine-preventable diseases, which can be used to inform public health programming.

The analysis then focuses on influenza, which accounts for a large proportion of reported vaccine-preventable disease cases and closely follows the category’s seasonal peaks. Trends in annual reported influenza cases and population-adjusted yearly rates are examined together to determine whether changes in reported illness persist after accounting for growth in Toronto’s population. 

These influenza patterns are then considered in the context of vaccination. Annual influenza vaccination is recommended by the Public Health Agency of Canada for most people aged six months or older, with particular emphasis on those who are at increased risk of severe illness [@phacFluShot]. National influenza vaccination coverage estimates provide additional context for interpreting the surveillance findings. Together, the data illustrate how disease surveillance and vaccination coverage estimates can be considered alongside one another to identify opportunities for targeted public health programming, without implying a direct causal relationship between vaccination rates and reported cases. 


# 2. Data

## 2.1 Toronto Communicable Disease Surveillance Data
Toronto Public Health (TPH) is responsible for updating and publishing a monthly dataset that describes reported cases of communicable diseases within the City of Toronto. These datasets are made available through the City of Toronto’s Open Data Portal under the title “Monthly Communicable Disease Surveillance Data” [@torontoCommunicableData].

As outlined by TPH, health surveillance programs, such as this one, help identify current public health issues. They also provide evidence to inform policy decisions aimed at reducing health risks and improving population health outcomes [@HSP_2026]. Similarly to TPH, Public Health Ontario (PHO) also provides information on reported disease cases for a variety of infectious diseases, filtered by region [@phoDiseaseTrends2026]. The PHO dataset serves as an alternative to the one used for this report and can be used to expand the analysis to investigate similar trends across multiple health units across Ontario.   

Reports from 2021 to 2026 were downloaded from Toronto's Open Data Portal using R [@r2026] and the `opendatatoronto` package [@opendatatoronto2025], with support from `dplyr` [@dplyr2026], `here` [@here2025] and `glue` [@glue2026]. The analysis includes complete calendar years from 2021 to 2025, with revised 2025 annual counts and rates taken from the 2026 report [@torontoCommunicableData]. All plots and tables were created using `ggplot2` [@ggplot22016] and `tinytable` [@tinytable2026], respectively, with support from `readr` [@readr2026].

`Python` was used to combine the reports into one dataset with a row for each disease and year [@python2026]. The rows correspond to reported case counts for a selected disease and year. Rows corresponding to COVID-19 were not included as they were out of scope of this report. The cleaned dataset consists of 331 rows, including 79 unique diseases across 6 distinct categories. An outline of the variables as well as their descriptions can be found in @tbl-variables and a sample of the dataset can be found in @tbl-sample. 


```{r}
#| label: tbl-variables
#| tbl-cap: "Fields found in the cleaned daatset with corresponding description."
#| echo: false

library(tinytable)

variable_descriptions <- data.frame(Variable = c("Disease", "Disease Category", "January-December", "Yearly Cases", "Yearly Rate", "Year"), 
                                    Description = c(
                                                    "Name of reported disease.", 
                                                    "The reporting category assigned to the disease.", 
                                                    "Reported case counts for the months of the calendar year.", 
                                                    "Reported case counts for the year of surveillance for a given disease.", 
                                                    "Rate of cases expressed as cases per 100,000 population for year.", 
                                                    "Year of observation."

                                                        )
                                                        )
tt(variable_descriptions)
```
## 2.2 Monthly Disease-Category Patterns
The monthly series of cases at the disease-category level provides a detailed view of how reported communicable diseases evolve between January 2021 and December 2025. For each category, monthly case totals are constructed by summing the reported cases of diseases assigned to that group. The resulting series therefore represents reported cases rather than unique individuals, since an individual may appear more than once across diseases. The diseases found within each disease category are summarized in @tbl-disease-categories. 

```{r}
#| label: tbl-disease-categories
#| tbl-cap: "Diseases included in each category, 2021–2025."
#| echo: false
#| warning: false

# Mark diseases that do not have an entry in all five study years.
diseases_by_category <- disease_data |>
  filter(Year %in% 2021:2025) |>
  group_by(`Disease Category`, Disease) |>
  summarise(Years = n_distinct(Year), .groups = "drop") |>
  mutate(Disease = if_else(Years < 5, paste0(Disease, "*"), Disease)) |>
  arrange(`Disease Category`, Disease) |>
  group_by(`Disease Category`) |>
  summarise(Diseases = paste(Disease, collapse = "; "), .groups = "drop")

tt(
  diseases_by_category,
  width = c(0.25, 0.75),
  notes = "* Not listed in every year from 2021 to 2025."
)
```

```{r}
#| label: prepare-monthly-data
#| include: false
#| echo: false
#| warning: false

# Stack the twelve month columns without reading the CSV again.
monthly_data <- bind_rows(lapply(seq_along(month.name), function(m) {
  disease_data |>
    transmute(Category = `Disease Category`, Disease, Year,
              Date = as.Date(sprintf("%d-%02d-01", Year, m)),
              Cases = .data[[month.name[m]]])
}))

# Sum only available counts and explicitly flag partial totals.
category_monthly <- monthly_data |>
  group_by(Category, Date) |>
  summarise(Cases = if (all(is.na(Cases))) 0 else sum(Cases, na.rm = TRUE),
            Incomplete = any(is.na(Cases)),
            .groups = "drop")

theme_set(theme_minimal(base_size = 10))
```

```{r}
#| label: fig-monthly-disease-categories
#| fig-cap: "Monthly reported cases by category, 2021-2025. Panels use different vertical scales."
#| echo: false 
#| warning: false
#| fig-width: 7
#| fig-height: 5.2

category_plot <- ggplot(category_monthly, aes(Date, Cases)) +
  geom_line(colour = "#235789", linewidth = 0.45) +
  geom_point(data = filter(category_monthly, Incomplete),
             shape = 1, size = 2, colour = "#A33A20") +
  facet_wrap(~Category, scales = "free_y", ncol = 2,
             labeller = label_wrap_gen(width = 36)) + 
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) format(x, big.mark = ",", scientific = FALSE)) +
  labs(title = "Monthly Cases by Disease Category", x = NULL, y = "Reported cases") +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 9))
category_plot
```

The monthly series shows substantial differences in both the volume and timing of reported cases across categories (@fig-monthly-disease-categories). Certain categories show a stronger seasonal structure. Most notably, vaccine-preventable cases are concentrated near the end and beginning of the calendar year, with distinct annual peaks occurring in December and January (December-2023, January-2024, December-2025). In contrast, enteric, food and waterborne diseases and vector-borne and zoonotic diseases rise during the summer months, with the latter peaking in July or August. The contrast shows why annual totals alone are insufficient to describe the timing of reported illness. 

Further understanding of the factors driving the seasonal trend can be gained by considering the contribution of individual diseases to the overall pattern within each category. The next section demonstrates an example of this analysis by examining influenza's contribution to the prominent seasonal pattern of vaccine-preventable diseases.


## 2.3 Influenza and Vaccine-Preventable Diseases
Influenza accounts for a significant proportion of the vaccine-preventable disease cases ranging from 81.4% to 94.1% between 2022 and 2025. @fig-influenza-vaccine-dep depicts this relationship wherein the monthly curve for influenza closely follows the overall category’s peaks. The low influenza count in 2021 is an exception, so the pattern should not be described as uniform across all five years. Comparing the two curves shows how much of the category's seasonal variation is associated with this single disease. 

```{r}
#| label: fig-influenza-vaccine-dep
#| fig-cap: "Monthly influenza cases and the vaccine-preventable category total, 2021-2025."
#| fig-width: 7
#| fig-height: 3.2
#| echo: false 
#| warning: false

comparison <- bind_rows(
  category_monthly |>
    filter(Category == "Vaccine Preventable Diseases") |>
    transmute(Date, Cases, Series = "Vaccine-preventable total"),
  monthly_data |>
    filter(Disease == "Influenza") |>
    transmute(Date, Cases, Series = "Influenza")
)

ggplot(comparison, aes(Date, Cases, colour = Series, linetype = Series)) +
  geom_line(linewidth = 0.65) +
  scale_colour_manual(values = c("Influenza" = "#C34A16",
                                "Vaccine-preventable total" = "#235789")) +
  scale_linetype_manual(values = c("Influenza" = "dashed",
                                  "Vaccine-preventable total" = "solid")) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(title = "Influenza and Vaccine-Preventable Cases",
       x = NULL, y = "Reported cases", colour = NULL, linetype = NULL) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The annual influenza rate declined to 105.8 per 100,000 in 2023, then increased to 148.3 in 2024 and 387.0 in 2025, the highest value observed. The increase from 2024 to 2025 is especially noticeable, showing that much of the recent rise occurred in the final year rather than building gradually across the whole period. This jump in population-adjusted rates suggests the increased case counts during seasonal peaks cannot be completely explained by population growth.

```{r}
#| label: fig-influenza-yearly-rate
#| fig-cap: "Annual reported influenza rates, 2021-2025."
#| fig-width: 7
#| fig-height: 3
#| echo: false 
#| warning: false

# Influenza has exact numeric annual rates. Do not turn missing rates into zeros.
influenza_rates <- disease_data |>
  filter(Disease == "Influenza") |>
  mutate(Rate = as.numeric(`Yearly Rate`))

ggplot(influenza_rates, aes(Year, Rate)) +
  geom_line(colour = "#235789", linewidth = 0.65) +
  geom_point(colour = "#235789", size = 2) +
  scale_x_continuous(breaks = 2021:2025) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = "Annual Influenza Rates",
       x = NULL, y = "Reported cases per 100,000 population", caption = "Note: Yearly rate of Infleunza in 2021 was 0.6%") +
  theme(panel.grid.minor = element_blank())
```

# 3 Discussion
## 3.1 Reported Influenze Cases and Vaccination Coverage
Influenza accounted for a large majority of vaccine-preventable disease cases between 2023 and 2025. The increase is also evident in population-adjusted incidence rates, indicating that the rise in reported cases cannot be explained by Toronto’s population growth alone. Together, these findings make influenza vaccination a particularly relevant focus for prevention, although vaccination coverage is not measured in the Toronto dataset [@torontoCommunicableData].

According to the Public Health Agency of Canada (PHAC), annual influenza vaccination is recommended for nearly everyone aged six months and older and reduces the risk of infection and flu-related complications. Vaccination also helps limit transmission to others, including those who are at greater risk of severe illness. PHAC identifies people aged 65 and older and adults aged 18-64 with chronic medical conditions as higher risk [@phacFluShot]. The national influenza vaccination coverage target is 80% for the two populations [@canada_srvcs_2026].

PHAC reports that vaccination among adults aged 18-64 without chronic medical conditions was 26.8%, 31.1%, 28.5% and 22.2% in the 2021-2022 through 2024-2025 seasons. In 2024-2025, coverage was 32.3% among adults aged 18-64 with chronic conditions and 63.2% among adults aged 65 and older [@canada_srvcs_2026].

For the 2025-2026 season, the survey found the main reasons for non-vaccination were the perception of good health, followed by “not getting round to it” [@canada_srvcs_2026]. This may point to a gap in community outreach, particularly in communicating the importance of annual influenza vaccination to people who are not considered high-risk. In 2025, the Minstry of Health Ontario issues an executive notice announcing that publicly funded influenza vaccines were available in pharmacies across the province [@ontarioInfluenza2025]. However, improving access alone may not be sufficient to increase vaccine uptake. Additional resources could therefore be directed toward reducing barriers to participation, including targeted communication campaigns and convenient vaccination clinics in workplaces and campuses.

Timing also matters. Influenza's largest spikes occurred in December 2023, January 2024 and December 2025. Protective antibody levels generally develop about two weeks after vaccination, and influenza vaccination is recommended annually [@phacInfluenzaGuide]. The practical message is to plan early in autumn and receive the vaccine when eligible and available, allowing time for an immune response before seasonal activity increases.

## 3.2 Limitations
A limitation of this analysis is its dependence on consistent reporting and testing for the diseases recorded. Additionally, national coverage estimates may not generalize to the population of Toronto and cannot show that low vaccination uptake caused the city's increase in influenza.

# Appendix 
## A. Sample Data 
```{r}
#| label: tbl-sample
#| tbl-cap: "Sample  entries from the cleaned dataset."


sample_rows <- disease_data |>
  slice(c(1, 10, 100, 250)) |>
  select(`Disease Category`, Disease, Year, January, December,
         `Yearly Cases`, `Yearly Rate`)
names(sample_rows) <- c("Category", "Disease", "Year", "Jan", "Dec", "Cases", "Rate")
tt(sample_rows, width = c(.18, .24, .08, .08, .08, .12, .12),
   notes = "Note: January and December illustrate the monthly fields. Rates are per 100,000 population.") |>
  style_tt(fontsize = 0.7)
```

# References
